Learning to Synthesize Rules for Rule-based Intrusion Detectors

### 2020-03-06 paper/paper.tex
handling comments

\item[\Contrib{}]\textbf{Data.} A data set of artifacts involved in
  the evaluation of \tname\ (\eg{}, pcap files of example attacks, list
  of plausible rules generated, etc.). The data set is available at
  the following URL~\url{https://zenodo.org/record/3698366}.

@@

\sloppy Network Intrusion Detection Systems (\nids{}) are typically categorized in two
groups~\cite{kumar2007survey}: rule-based, as briefly described on
Section~\ref{sec:intro}, and anomaly-based \nids. An
\emph{anomaly-based intrusion detector} learns usage behavior from
benign traffic and, based on that, alerts uncommon
behavior~\cite{7579764,kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}. More
precisely, an anomaly-based \nids{} flags an attack when the network
traffic manifests different characteristics compared to those of the
benign traffic. Rule-based \nids{} and anomaly-based \nids{} are
complementary. A rule-based \nids{} focuses on known attacks whereas
anomaly-based \nids{} focuses on unknown potential attacks. Rule-based
\nids\ can miss attacks as rulesets can become obsolete whereas
anomaly-based \nids\ can report false alarms as learning algorithms
that generalize observations from benign traffic are approximate.
This paper focuses on rule-based \nids\ for its tremendous popularity
in industry\Fix{cite or tone down}. Although we used Suricata, the
principles we used are general to any other signature-based
\nids~(\eg{}, Snort~\cite{snort}).

@@

\vspace{-1ex}
\begin{table}[t!]
  \footnotesize
  \caption{\label{table:rules}Example of non-documentation options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{6cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the frag. or reserved bits are set in the IP header\\
    9 & threshold & minimum threshold for a rule before it generates alerts\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
  \vspace{-3ex}  
\end{table}

@@

We mentioned above that http rules often used content options, which
associate with the message payload. For non-http rules, options are
often associated with packet fields. Let us consider a rule for an
icmp attack as example. Figure~\ref{fig:icmp-packet-layout} shows the
layout of an icmp packet. The packet fields ``Flags'' and ``Time to
Live'', declared in the ip header, are mapped to the \suri\ options
\CodeIn{flagbits} and \CodeIn{ttl}, respectively. Likewise, the fields
``Type'' and ``Code'', declared in the icmp header, are mapped to the
options \CodeIn{itype} and \CodeIn{icode}, respectively. The values of
the options ``ttl'', ``itype'', and ``icode'' are the same as the
value of the corresponding fields in the icmp packet. For the field
``Flags'' the value of the corresponding option is encoded in a
bitvector. For example, if the ``Flags'' field stores the value 1010,
the corresponding rule option will be \CodeIn{flagbits:RM}, as each
position of the vector encodes a different character in the
\CodeIn{flagbits} rule option. In summary, the value of several
options can be determined through transformation functions, as those
informally described above. One exception to this is the option
\CodeIn{threshold}, which is typically used in denial-of-service
attacks. That option operates over sequence of messages as opposed to
a single message. Consequently, it is not possible to determine its
values from a single message.

\begin{figure}[t!]
\centering
\includegraphics[scale=0.27]{figs/ICMP-packet-structure.png}
\vspace{-2ex}
\caption{Layout of an icmp packet.}
\vspace{-2ex}
\label{fig:icmp-packet-layout}
\end{figure}
%\vspace{-2ex}

@@

%\vspace{-1ex}
\begin{figure}[h!]
  \lstinputlisting[language=C,numbers=none,keywords={dsize,itype}]{pingscan.suricata}
  \vspace{-2ex}  
  \caption{Suricata rule for Ping Scan attack.}
  \label{fig:pingscan-example}
\end{figure}
\vspace{-3ex}

@@

To answer this research question, we selected two cases with the
highest number of examples (see column ``\#~Vars'' on
Table~\ref{table:results})---idqaccess and scripttag---and the case
where \tname\ performs worst---jboss \Fix{Remover jboss}. We measured the position of the
golden rule at the rank for an increasing number of of examples
provided to \tname. For each number of variants, we sampled five sets
of that size to provide as input to the
technique. Figure~\ref{fig:impact-number-attacks} shows the results
for each attack we selected. The y-axis of each plot denotes the rank
of the golden rule and x-axis denotes the number of attack examples
used to synthesize rules. Each point in the x-axis is associated with
the boxplot of the distribution of results for the corresponding
number of variants.

@@

Results indicate that 1) the number of examples is important, 2) the
choice of examples is important, and 3) a small number of examples
suffices to reach the best results.

@@

\begin{figure}[H]
  \lstinputlisting[language=C,numbers=none,keywords={content,flow}]{adaptor-golden-rule.suricata}
  \vspace{-2ex}  
  \caption{\label{fig:rule-jboss}Suricata rule for JBoss Remote Code
    Execution.}
  \vspace{-3ex}
\end{figure}

@@

Figure~\ref{fig:rule-jboss} shows a Suricata rule that captures a
JBoss RCE attack.  The option \CodeIn{flow: established, to\_server;}
indicates that a rule match will only occur on packets that (1) flow
on an already-established connection and that (2) flow to the server
as opposed to the client. For the content options the modifiers
\CodeIn{nocase} and \CodeIn{http\_uri} indicate, respectively, that
the matching is not case sensitive and that the string should be
located at the packet field ``Request URI''. We reproduced the
malicious JBoss attack using the traffic from a Nikto
script~\cite{nikto}. Similar attacks are also available on the
Metasploit database~\cite{metasploit}. This is a case where
\tname\ performed poorly. Of the 228 plausible rules it generated for
this attack, the ground truth was located at position 45 in the ranked
list. The reason for that is that \Fix{...xxx...}

%% \Mar{move from above --> We used the following hping3 command to
%%   reproduce that attack: \CodeIn{hping3 -d 80 -w 512 -S -p 80 --flood
%%     --rand-source <IP>}.}

@@

\noindent
\textbf{Alternative methods for rule synthesis.} We discuss in the
following techniques similar in goal to ours. Unfortunately, we could
not access the tools implementing the techniques for side
comparison. Nevertheless, we found that there are important
differences between our work and related work. Vollmer
\etal{}~\cite{vollmer-etal-cics2011} propose the use of search to
derive rules for Snort, which are very similar to \suri's.  Their
approach focuses on icmp attacks. It models every field of the icmp
packet in the chromosome of candidate solutions (\ie{}, a rule) of a
genetic search. The initial population is a set of candidate solutions
that randomly combine options, model-ling the packet fields. They show
that they were able to generate rules without false negatives and only
3 false positives across 33K benign packets. The key differences
between our work and Vollmer~\etal{}'s are 1) the support of different
protocols and 2) the use of ranking to discriminate alternative
plausible rules. As for 1, our approach supports a variety of
protocols and options that refer to message payloads which
significantly increase the search space. Note that icmp rules only
refer to message fields (see Section~\ref{sec:rules-and-packets})
\Luc{eh possivel incluir contents em regras icmp tb. o q a secao 2.3 
diz q \textbf{geralmente} as regras nao-http possuem apenas field options}. 
As for 2, our solution additionally ranks plausible options, which
contain \emph{no} false positives or negatives with respect to the
benign and mallicious traffic considered. Muthuregunathan
\etal{}~\cite{muthuregunathan-etal2009} propose a clustering method to
create rules. They characterize candidate solutions as a set of field
values and use a clustering algorithm---hybrid k-means partitioned
around menoids---to isolate malicious from benign
traffic. Subsequently, they optimize the formed clusters using genetic
algorithms and hill-climbing search. Ou approach differs from theirs
in the method used, results, and support for content
options. Kao~\etal~\cite{Kao2015AutomaticNR} proposed ARGS, a system
to synthesize http rules that take malicious traffic as input. ARGS
looks for a common pattern across the different inputs to define
rules. It does not take existing rules as input or benign
traffic. Consequently, false positive can be produced and need to be
handled manually by users. Our approach, in contrast, does not
generate false positives w.r.t. the input traffic.

@@

\noindent
\textbf{Anomaly-based intrusion detection.}
Cordy~\etal{}\cite{cordy-etal-issta19} propose the use of
\Fix{...what...} to \Fix{...calibrate parameters of?....}
anomaly-based \nids. Although our approach and theirs use machine
intelligence to detect attacks, as discussed on
Section~\ref{sec:background} anomaly-based and rule-based \nids\ are
complementary.
\Gui{Anomaly-based intrusion detection. Cordy and colleagues[28] 
propose the use of a search-based genetic algorithm to not only 
generate attack schemes capable of penetrating the tested anomaly-based 
NIDS, but also repairing the NIDS to identify the given attack scheme. 
Although our approach and theirs use machine intelligence to detect 
attacks, as discussed on Section 2 anomaly-based and rule-based NIDS 
are complementary.}

@@

Network Intrusion Detectors are an important defense mechanism against
networks attacks. This paper presents \tname{}, a novel approach to
generate rules for Rule-based \nids. \tname\ leverages positive and
negative traffic and examples of trusted rules to synthesize
rules. \tname{} is organized as a pipeline of three components that
1)~create an over-specified seed rule, 2)~derive plausible rules from
the seed, and 3)~rank plausible rules. Experimental results provide
early evidence showing that \tname\ is very promising. It was able to
rank the correct rule among the first five rules of the reported
ranking in \percTopFiveRanking\ of the cases we analyzed.

>>>>

\item[\Contrib{}]\textbf{Data.} A data set of artifacts involved in
  the evaluation of \tname\ (\eg{}, pcap files of example attacks, list
  of plausible rules generated, etc.). The data set is available at
  the following URL~\ourdataset.

@@

\sloppy Network Intrusion Detection Systems (\nids{}) are typically categorized in two
groups~\cite{kumar2007survey}: rule-based, as briefly described on
Section~\ref{sec:intro}, and anomaly-based \nids. An
\emph{anomaly-based intrusion detector} learns usage behavior from
benign traffic and, based on that, alerts uncommon
behavior~\cite{7579764,kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}. More
precisely, an anomaly-based \nids{} flags an attack when the network
traffic manifests different characteristics compared to those of the
benign traffic. Rule-based \nids{} and anomaly-based \nids{} are
complementary. A rule-based \nids{} focuses on known attacks whereas
anomaly-based \nids{} focuses on unknown potential attacks. Rule-based
\nids\ can miss attacks as rulesets can become obsolete whereas
anomaly-based \nids\ can report false alarms as learning algorithms
that generalize observations from benign traffic are approximate.
This paper focuses on rule-based \nids\ for its popularity
in industry. Although we used Suricata, the
principles we used are general to any other signature-based
\nids~(\eg{}, Snort~\cite{snort}).

@@

\vspace{-1ex}
\begin{table}[t!]
  \footnotesize
  \caption{\label{table:rules}Example of non-documentation options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{6cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the frag. or reserved bits are set in the IP header\\
    9 & threshold & minimum threshold for a rule before it generates alerts\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
%  \vspace{-3ex}  
\end{table}

@@

We mentioned above that http rules often used content options, which
associate with the message payload. For non-http rules, options are
often associated with packet fields. Let us consider a rule for an
icmp attack as example.

\begin{figure}[h!]
  \centering
  \vspace{-2ex}  
  \includegraphics[scale=0.27]{figs/ICMP-packet-structure.png}
  \vspace{-2ex}
  \caption{Layout of an icmp packet.}
  \vspace{-2ex}
  \label{fig:icmp-packet-layout}
\end{figure}

Figure~\ref{fig:icmp-packet-layout} shows the
layout of an icmp packet. The packet fields ``Flags'' and ``Time to
Live'', declared in the ip header, are mapped to the \suri\ options
\CodeIn{flagbits} and \CodeIn{ttl}, respectively. Likewise, the fields
``Type'' and ``Code'', declared in the icmp header, are mapped to the
options \CodeIn{itype} and \CodeIn{icode}, respectively. The values of
the options ``ttl'', ``itype'', and ``icode'' are the same as the
value of the corresponding fields in the icmp packet. For the field
``Flags'' the value of the corresponding option is encoded in a
bitvector. For example, if the ``Flags'' field stores the value 1010,
the corresponding rule option will be \CodeIn{flagbits:RM}, as each
position of the vector encodes a different character in the
\CodeIn{flagbits} rule option. In summary, the value of several
options can be determined through transformation functions, as those
informally described above. One exception to this is the option
\CodeIn{threshold}, which is typically used in denial-of-service
attacks. That option operates over sequence of messages as opposed to
single messages.

%Consequently, it is not possible to determine its
%values from a single message.

@@

%\vspace{-1ex}
\begin{figure}[h!]
  \lstinputlisting[language=C,numbers=none,keywords={dsize,itype}]{pingscan.suricata}
  \vspace{-2ex}  
  \caption{Suricata rule for Ping Scan attack.}
  \vspace{-4ex}  
  \label{fig:pingscan-example}
\end{figure}

@@

%%---and the case where \tname\ performs worst---jboss \Fix{Remover jboss}

To answer this research question, we selected two cases with the
highest number of examples (see column ``\#~Vars'' on
Table~\ref{table:results})---idqaccess and scripttag. We measured the
position of the golden rule at the rank for an increasing number of
examples provided to \tname. For each number of variants, we sampled
five sets of that size to provide as input to the
technique. Figure~\ref{fig:impact-number-attacks} shows the results
for each attack we selected. The y-axis of each plot denotes the rank
of the golden rule and x-axis denotes the number of attack examples
used to synthesize rules. Each point in the x-axis is associated with
the boxplot of the distribution of results for the corresponding
number of variants. Results indicate that 1) the number of examples is
important, 2) the choice of examples is important, and 3) a small
number of examples suffices to reach the best results.

@@

@@

@@

Figure~\ref{fig:rule-jboss} shows a Suricata rule that captures a
JBoss RCE attack.  The option \CodeIn{flow: established, to\_server;}
indicates that a rule match will only occur on packets that (1) flow
on an already-established connection and that (2) flow to the server
as opposed to the client. For the content options the modifiers
\CodeIn{nocase} and \CodeIn{http\_uri} indicate, respectively, that
the matching is not case sensitive and that the string should be
located at the packet field ``Request URI''. We reproduced the
malicious JBoss attack using the traffic from a Nikto
script~\cite{nikto}. Similar attacks are also available on the
Metasploit database~\cite{metasploit}. This is a case where
\tname\ performed poorly. Of the 228 plausible rules it generated for
this attack, the ground truth was located at position 45 in the ranked
list.

\begin{figure}[t!]
%  \vspace{-1ex}    
  \lstinputlisting[language=C,numbers=none,keywords={content,flow}]{adaptor-golden-rule.suricata}
  \vspace{-2ex}  
  \caption{\label{fig:rule-jboss}Suricata rule for JBoss Remote Code
    Execution.}
%  \vspace{-1ex}
\end{figure}

@@

\noindent
\textbf{Alternative methods for rule synthesis.} We discuss in the
following techniques similar in goal to ours. Unfortunately, we could
not access the tools implementing the techniques for side
comparison. Nevertheless, we found that there are important
differences between our work and related work. Vollmer
\etal{}~\cite{vollmer-etal-cics2011} propose the use of search to
derive rules for Snort, which are very similar to \suri's.  Their
approach focuses on icmp attacks. It models every field of the icmp
packet in the chromosome of candidate solutions (\ie{}, a rule) of a
genetic search. The initial population is a set of candidate solutions
that randomly combine options, modelling the packet fields. They show
that they were able to generate rules without false negatives and only
3 false positives across 33K benign packets. The key differences
between our work and Vollmer~\etal{}'s are 1) the support of different
protocols and 2) the use of ranking to discriminate alternative
plausible rules. As for 1, our approach supports a variety of
protocols and options that refer to message payloads which
significantly increase the search space. The icmp rules they generate only
refer to message fields (see Section~\ref{sec:rules-and-packets}).
\MyComment{\Luc{eh possivel incluir contents em regras icmp tb. o q a secao 2.3 
diz q \textbf{geralmente} as regras nao-http possuem apenas field options}. }
As for 2, our solution additionally ranks plausible options, which
contain \emph{no} false positives or negatives with respect to the
benign and mallicious traffic considered. Muthuregunathan
\etal{}~\cite{muthuregunathan-etal2009} propose a clustering method to
create rules. They characterize candidate solutions as a set of field
values and use a clustering algorithm---hybrid k-means partitioned
around menoids---to isolate malicious from benign
traffic. Subsequently, they optimize the formed clusters using genetic
algorithms and hill-climbing search. Ou approach differs from theirs
in the method used, results, and support for content
options. Kao~\etal~\cite{Kao2015AutomaticNR} proposed ARGS, a system
to synthesize http rules that take malicious traffic as input. ARGS
looks for a common pattern across the different inputs to define
rules. It does not take existing rules as input or benign
traffic. Consequently, false positive can be produced and need to be
handled manually by users. Our approach, in contrast, does not
generate false positives w.r.t. the input traffic.

@@

\noindent
\textbf{Anomaly-based intrusion detection.}
Cordy~\etal{}\cite{cordy-etal-issta19} propose the use of a genetic
algorithm to not only generate attack schemes capable of penetrating
the tested anomaly-based NIDS, but also repairing the NIDS to identify
the given attack scheme. Although our approach and theirs use machine
intelligence to detect attacks, anomaly-based and rule-based
\nids\ are complementary, as discussed on
Section~\ref{sec:background}.

@@

Network Intrusion Detectors are an important defense mechanism against
networks attacks. This paper presents \tname{}, a novel approach to
generate rules for Rule-based \nids. \tname\ leverages positive and
negative traffic and examples of trusted rules to synthesize
rules. \tname{} is organized as a pipeline of three components that
1)~create an over-specified seed rule, 2)~derive plausible rules from
the seed, and 3)~rank plausible rules. Experimental results provide
early evidence showing that \tname\ is very promising. It was able to
rank the correct rule among the first five rules of the reported
ranking in \percTopFiveRanking\ of the cases we analyzed. The data set
containing our experimental artifacts is publicly accessible at \ourdataset.
